---
title: " Real data: CAPE Positive"
subtitle:  "In: A guide to test association between Polygenic Risk Scores and psychological and psychiatric traits: practical examples"
author: "Itziar Irigoien, Patricia Mas, Sergi Papiol, Neus Barrantes-Vidal, Araceli Rosa, and Concepción Arenas"
output:
  pdf_document:
    keep_tex: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
source("Functions.R")
library(ggplot2)
library(car)
library(multcomp)
library(lattice)
```


## Working flow and code

In this real data set there are 106 PRS, and a **binary Trait** ($CAPE_{Positive}$), with gender, age, and two Principal Components as covariates. 

- Data reading
```{r}
dat <- read.table("Real_data_Positive.csv", header=TRUE, sep="\t", dec=".")
names(dat) #
dat <- dat[, -1]
```
Check that all variables you are interested in are properly read and that there are not other variables you do not need.

\bigskip

- Do not forget to declare the categorical variables as factors
```{r}
dat$Sex <- as.factor(dat$Sex)
dat$CAPE_Positive <- as.factor(dat$CAPE_Positive)
```

## 1.	What full model should be considered?

First, given a particular PRS (named PRS.i), consider all the possible full models:

- FM$_{WI}$ : log(p/1-p) versus PRS.i + Sex + Age + PC1 + PC2
- FM$_{Sex}$: log(p/1-p) versus PRS.i + Sex + PRS.i\,Sex + Age + PC1 + PC2


## 2. How to make a PRS ranking to find the important ones?

As is described in the paper, for each model, calculate the Tjur's coefficients of discrimination. If Nagelkerke's $R^2$ is prefered, set statistic="PseudoR2" in function orderBin(), and calculate their sum $S$. 

According to S, list the PRSs in decreasing order:
```{r}
# Order the PRSs
out <- orderBin(dat, yname="CAPE_Positive", prsname = "PRS.", statistic = "D")
head(out)
mainfilename <- "Real_data_CAPE_Positive"
filename <- paste0(mainfilename, "_Ordered_PRS.csv")
write.csv2(out,file=filename)
```

Plot the sum of discrimination coefficients  $D$. Lines: in blue the median; in black the mean.
```{r}
out <- data.frame(out) 
n <- dim(out)[1]
select <- grep("Model", names(out), value=FALSE)
out$effect <- out$Sum
sds <- apply(out[, select], 1, sd)
out$lower <- out$effect - sds
out$upper <- out$effect + sds
out$rank <- n:1


ggplot(data=out, aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum of D', y = 'PRS') +
  geom_vline(xintercept=mean(out$effect), color='black', linetype='dashed') +
  geom_vline(xintercept=median(out$effect), color='blue', linetype='dashed') +
  theme_minimal()

```

According to the obtained results, first PRS.15 is selected to analyse its association with the $CAPE_{Positive}$.

### 3. Which model, of all the possible ones, should be used?

The following Figure represents the scatter plot separated by Sex group.
```{r}
# First candidate PRS.15
# Plot it
M <- glm(CAPE_Positive ~ PRS.15*Sex + Age + PC1 + PC2, data=dat, family=binomial())
pre <- M$fitted.values #predict(M,type='response')   
xyplot(log(pre/(pre+1))~PRS.15|Sex, data=dat,  type=c("p", "r"))
```

The plots suggest that the interaction between the PRS.15 and the diagnostic is relevant. Thus, we set the full model candidate (FM): $CAPE_{Positive} \sim PRS + Sex  + PRS\cdot Sex + Age + PC1 +PC2$

### 5.	For a binary trait, what steps should be followed for a correct analysis?

Check for overdispersion
```{r}
#model
FM <- glm(CAPE_Positive ~ Sex + PRS.15*Sex + Age + PC1 + PC2, data=dat, family=binomial())
#Residual Deviance
FM$deviance
# Ratio 
FM$deviance/FM$df.residual

```

Since this ratio is close to 1, there is not evidence of overdispersion.

```{r}
#With chi-squared test
FM.od <- glm(CAPE_Positive ~ Sex + PRS.15*Sex + Age + PC1 + PC2, data=dat, family=quasibinomial())
pchisq(summary(FM.od)$dispersion * FM$df.residual,
       FM$df.residual, lower = F)
```

With this p-value = 0.2651 we conclude that there is not evedence of overdispersion.

Based on the estimations given in this table:
```{r}
summary(FM)
```

The PRS.15 coefficient will vary depending on the gender, being:

  - $\widehat{log(p/1-p)} = 0.055 + 0.746\times PRS.15 - 0.118\times Age + 1.951\times PC1 + 2.867\times PC2$, if Sex = 0.
  - $\widehat{log(p/1-p)} = (0.055 + 0.139) + (0.746 - 0.749)\times PRS.15 - 0.118\times Age + 1.951\times PC1 + 2.867\times PC2$, if Sex = 1.



Check whether the respective PRS coefficients under each group are significant or not.
```{r}
summary(glht(FM, "PRS.15 = 0"))
summary(glht(FM, "PRS.15  + Sex1:PRS.15 = 0"))
```

That means that for those with Sex=1, it seems that the PRS.15 is not related to the Trait with odds = exp(-0.002) = 0.991, but for those with Sex = 0 the model indicates that the coefficient of PRS.15 is 0.7463, so the odds increase exp(0.7463) = 2.109 for an incremental of one unit in PRS.15 with a p-value=0.0101.


It is possible to compute a permutation test to assess whether the increase in the coefficient of determination $D$ is significative.
```{r}
# Null model
NM <- glm(CAPE_Positive ~  Sex + Age + PC1 + PC2, data=dat, family=binomial() )
permtest <- dD(NM, FM, seed=1236)
permtest
```
In this particular case, it can be seen that the coefficient of discrimination of the FM model is 0.05 units bigger than the corresponding to the *Null Model* NM, and it is statistically significant.

- **Last step: We move to the next PRS.**

