
- **Title: Real data CAPE Negative**
- **Output: html_document**
- **Reference: Tutorial: a guide to test evidence of association between a Polygenic Risk Score and a Trait**
- **Authors: Itziar Irigoien, Patricia Mas, Araceli Rosa, and Concepción Arenas**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
source("Functions.R")
library(ggplot2)
library(car)
library(multcomp)
```


## **Real data CAPE Negative**

- **In this real data set there are 106 PRS, and a continuous Trait (CAPE_Negative), with gender, age, and two Principal Components as covariates.**

- **data reading**
```{r}
dat <- read.table("Real_data.csv", header=TRUE, sep=";", dec=",")
names(dat) #
```

- **do not forget to declare the categorical variables as factors**
```{r}
dat$Sex <- as.factor(dat$Sex)
```

- **Step 1: Order  the PRSs**
```{r}
out <- orderR2(dat, yname="CAPE_Negative", prsname = "PRS.")
head(out)
mainfilename <- "Real_example_CAPE_Negative"
filename <- paste0(mainfilename, "_Ordered_PRS.csv")
write.csv2(out,file=filename)
```

- **Step 1: Plot the sum of coefficients of determination $S_{R^2}$. Lines: in blue the median; in black the mean**
```{r}
out <- data.frame(out) 
nPRS <- dim(out)[1]
select <- grep("Model", names(out), value=FALSE)
out$effect <- out$Sum
sds <- apply(out[, select], 1, sd)
out$lower <- out$effect - sds
out$upper <- out$effect + sds
out$rank <- nPRS:1

n <- dim(out)[1]
ggplot(data=out, aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=mean(out$effect), color='black', linetype='dashed') +
  geom_vline(xintercept=median(out$effect), color='blue', linetype='dashed') +
  theme_minimal()

```

- **Repeat the plot with only de PRS with $S_{R^2}$ larger then the median. Lines: in red and solid the 95% percentile; in red and dashed the 90%, 85%, 80% percentiles**
```{r}
sel <- out$Sum > median(out$effect)
ggplot(data=out[sel,], aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=quantile(out$Sum, prob=0.95), color='red', linetype='solid') +
  geom_vline(xintercept=quantile(out$Sum, prob=c(0.80, 0.85, 0.90)), color='red', linetype='dashed') +
  theme_minimal()
```

- **Create files with PRS with the sum of $R^2$ greater then the 95%, 90%, 85% and 80% percentile**
```{r}
sel_95 <- out$Sum > quantile(out$Sum, prob=0.95)
write.csv2(out[sel_95,-9], file="CAPE_Negative_percentil_95.csv")
sel_90 <- out$Sum > quantile(out$Sum, prob=0.90)
write.csv2(out[sel_90,-9], file="CAPE_Negative_percentil_90.csv")
sel_85 <- out$Sum > quantile(out$Sum, prob=0.85)
write.csv2(out[sel_85,-9], file="CAPE_Negativ_85.csv")
sel_80 <- out$Sum > quantile(out$Sum, prob=0.80)
write.csv2(out[sel_80,-9], file="CAPE_Negativ_80.csv")

```

- **According to the obtained results, first, PRS.13 is selected to analyse its possible association with the CAPE_Negative**

- **Step 2.1: Plots to check the possible interactions**
```{r}
# First candidate PRS.13
# Plot it
library(lattice) 
xyplot(CAPE_Negative~PRS.13|Sex, data=dat,  type=c("p", "r"))
```

- **The plots suggest that the interaction between the PRS.13 and the sex is not relevant. Thus, we set the full model candidate (FM): $CAPENegative \sim PRS + Sex + Age + PC1 +PC2$**


- **Step2.2: Do parametric assumptions for the selected FM hold?**
```{r}
#model
FM <- lm(CAPE_Negative ~ PRS.13 + Sex + Age + PC1 + PC2, data=dat)
#qq-plot for normality 
plot(FM,2)   
```

```{r}
#Saphiro-Wilk test
shapiro.test(FM$residuals)
```

```{r}
#plot for variances
d <- fortify(FM)
ggplot(d,aes(x=.fitted, y=.stdresid, colour=Sex)) + geom_point() + geom_hline(yintercept=0, col="red")+
  facet_wrap(.~Sex)
```

```{r}
#Levene's test
leveneTest(.stdresid ~ Sex, data=d)
```

- **It seems that there is lack of normality of residual but linearity and homocedasticity assumptions hold. Thus, we have two approaches to assess the possible association with PRS.13 and the Trait: try a transformation (**Step 4.1.1**) or perform a permutation test (**Step 4.1.2**)**

- **Step 4.1.1: Try a transformation of the dependent variable:**
```{r}
dat$TNegative <- sqrt(dat$CAPE_Negative)
FM <- lm(TNegative ~ PRS.13 + Sex + Age + PC1 + PC2, data=dat)
#qq-plot for normality 
plot(FM,2)
```

```{r}
#Shapiro-Wilk test
shapiro.test(FM$residuals)
```

- **Since it is suggested the transformation offered a solution for the lack of normality, the model we build is given by:**
```{r}
summary(FM)
```

- **The results show that PRS.13 is related with the sqrt(Trait) in the following way:**

\begin{itemize}
    \item $\widehat{sqrt(Trait)} = 3.013 - 0.101\times PRS.13 - 0.142\times Sex + 0.009\times Age + 3.797\times PC1 + 8.409\times PC2$ 

\end{itemize}

- **Step 4.1.2: Perform a permutation test to assess the association**
```{r}
NM <- lm(CAPE_Negative ~  Sex + Age + PC1 + PC2, data=dat)
FM <- lm(CAPE_Negative ~ PRS.13 + Sex + Age + PC1 + PC2, data=dat)
outperm <- dR2(NullModel=NM, FullModel=FM, B=5000, seed=165)
outperm
```
- ** We observe an increase of 0.0196 in the coefficient of determination when the PRS.13 is included in the model and the permutation test indicates it is significant.**


- **Last step: We move to the next PRS**

