---
title: "Real data: CAPE Negative"
subtitle:  "In: A guide to test association between Polygenic Risk Scores and psychological and psychiatric traits: practical examples"
author: 
  - "Itziar Irigoien, Patricia Mas-Bermejo, Sergi Papiol, Neus Barrantes-Vidal,"
  - "Araceli Rosa, and Concepción Arenas"
output:
  pdf_document:
    keep_tex: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
source("Functions.R")
library(ggplot2)
library(car)
library(multcomp)
```


## Working flow and code

In this real data set there are 106 PRS, and a **continuous Trait** ($CAPE_{Negative}$), with sex, age, and two Principal Components as covariates. For more details see Section 7 in the paper.

- Data reading
```{r}
dat <- read.table("Real_data_Negative.csv", header=TRUE, sep="\t", dec=".")
names(dat) #
dat <- dat[, -1]
```
Important! Check that all variables you are interested in are properly read and that there are not other variables you do not need.

\bigskip

- Do not forget to declare the categorical variables as factors.
```{r}
dat$Sex <- as.factor(dat$Sex)
```

## 1.	What full model should be considered?

First, given a particular PRS (named PRS.i), consider all the possible full models:

- FM$_{WI}$ : Trait versus PRS.i + Sex + Age + PC1 + PC2
- FM$_{Sex}$: Trait versus PRS.i + Sex + PRS.i · Sex +  Age + PC1 + PC2


## 2. How to make a PRS ranking to find the important ones?

As is described in the paper, for each model, calculate the coefficient of determination $R^2$ and calculate the sum: $S =  R^2_{WI} +  R^2_{Sex}$. 

According to S, list the PRSs in decreasing order:
```{r}
out <- orderR2(dat, yname="CAPE_Negative", prsname = "PRS.")
head(out)
mainfilename <- "Real_example_CAPE_Negative"
filename <- paste0(mainfilename, "_Ordered_PRS.csv")
write.csv2(out,file=filename)
```

Plot the sum of coefficients of determination $S_{R^2}$. Lines: in blue the median; in black the mean.
```{r}
out <- data.frame(out) 
nPRS <- dim(out)[1]
select <- grep("Model", names(out), value=FALSE)
out$effect <- out$Sum
sds <- apply(out[, select], 1, sd)
out$lower <- out$effect - sds
out$upper <- out$effect + sds
out$rank <- nPRS:1

n <- dim(out)[1]
ggplot(data=out, aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=mean(out$effect), color='black', linetype='dashed') +
  geom_vline(xintercept=median(out$effect), color='blue', linetype='dashed') +
  theme_minimal()

```


According to the obtained results, first, PRS.13 is selected to analyse its possible association with the $CAPE_{Negative}$.

## 3. Which model, of all the possible ones, should be used?

The following Figure represents the scatter plot of $CAPE_{Negative}$ versus PRS.13 separated by Sex group.
```{r}
# First candidate PRS.13
# Plot it
library(lattice) 
xyplot(CAPE_Negative~PRS.13|Sex, data=dat,  type=c("p", "r"))
```

The plots suggest that the interaction between the PRS.13 and the sex is not relevant. Thus, we set the full model candidate (FM): $CAPE_{Negative} \sim PRS + Sex + Age + PC1 +PC2$.

## 4.	For a continuous trait, what steps should be followed for a correct analysis?

- **4.1.  How is the candidate model validated?**

First, we validate the normality of the errors and the constant variance conditions (see the figures and the results of Shapiro test and Levene test).

```{r}
#model
FM <- lm(CAPE_Negative ~ PRS.13 + Sex + Age + PC1 + PC2, data=dat)
#qq-plot for normality 
plot(FM,2)   
```

The lack of normality of residuals is suggested by this last plot.

This supported by Shapiro’s test:
```{r}
#Shapiro-Wilk test
shapiro.test(FM$residuals)
```

```{r}
#plot for variances
d <- fortify(FM)
ggplot(d,aes(x=.fitted, y=.stdresid, colour=Sex)) +
  geom_point() +
  geom_hline(yintercept=0, col="red")+
  facet_wrap(.~Sex)
```

```{r}
#Levene's test
leveneTest(.stdresid ~ Sex, data=d)
```

All in all, it seems that there is lack of normality of residuals but linearity and homocedasticity assumptions hold.


- **4.2. What can be done if any validation condition fails?**

We have two approaches to assess the possible association with PRS.13 and $CAPE_{Negative}$: try a transformation or perform a permutation test.


First we try the squared root transformation for the dependent variable:
```{r}
dat$TNegative <- sqrt(dat$CAPE_Negative)
FM <- lm(TNegative ~ PRS.13 + Sex + Age + PC1 + PC2, data=dat)
#qq-plot for normality 
plot(FM,2)
```

```{r}
#Shapiro-Wilk test
shapiro.test(FM$residuals)
```
It is suggested the transformation offered a solution for the lack of normality. Furthermore,...
```{r}
#plot for variances
d <- fortify(FM)
ggplot(d,aes(x=.fitted, y=.stdresid, colour=Sex)) +
  geom_point() + 
  geom_hline(yintercept=0, col="red")+
  facet_wrap(.~Sex)
```

```{r}
#Levene's test
leveneTest(.stdresid ~ Sex, data=d)
```

...results do not indicate evidence against homoscedasticity, neither a pattern is observed that could indicate a lack of linearity.

Therefore, the model we build is given by:
```{r}
summary(FM)
```

The results show that PRS.13 is related with the $\sqrt{Trait}$ in the following way:

$$\widehat{\sqrt{Trait}} = 3.013 - 0.101\times PRS.13 - 0.142\times Sex + 0.009\times Age + 3.797\times PC1 + 8.409\times PC2,$$ 
where Sex takes values 0 or 1, depending on whether the individual under study is male or female. See section 7.1 in the paper for more details.


On the other hand, based on the permutation approach:
```{r}
NM <- lm(CAPE_Negative ~  Sex + Age + PC1 + PC2, data=dat)
FM <- lm(CAPE_Negative ~ PRS.13 + Sex + Age + PC1 + PC2, data=dat)
outperm <- dR2(NullModel=NM, FullModel=FM, B=5000, seed=165)
outperm
```
We observe an increase of 0.0196 in the coefficient of determination when the PRS.13 is included in the model and the permutation test indicates it is significant.


- **Last step: We move to the next PRS.**

