
- **Title: Working example 3**
- **Output: html_document**
- **Reference: Tutorial: a guide to test evidence of association between a Polygenic Risk Score and a trait**
- **Authors: Itziar Irigoien, Patricia Mas, Araceli Rosa, and Concepción Arenas**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
source("Functions.R")
library(ggplot2)
library(car)
library(MASS)
```


## Working example 3

## In this example we simulate 5 PRSs, and a continuous trait, with gender, clinical diagnosis (with 2 categories), age, and two Principal Components as covariates. 

- **data reading**
```{r}
dat <- read.table("WExample3.csv", header=TRUE, sep=";", dec=",")
names(dat) #
```

- **do not forget to declare the categorical variables as factors**
```{r}
dat$Sex <- as.factor(dat$Sex)
dat$Diagnostic <- as.factor(dat$Diagnostic)
```

- **Step 1: Order  the PRSs**
```{r}
out <- orderR2(dat, yname="Trait", prsname = "PRS.")
head(out)
write.csv2(out,file="WExample1_Ordered_PRS.csv")
```

- **Step 1: Plot the sum of coefficients of determination $S_{R^2}$. Lines: in blue the median; in black the mean**
```{r}
out <- data.frame(out) 
nPRS <- dim(out)[1]
select <- grep("Model", names(out), value=FALSE)
out$effect <- out$Sum
sds <- apply(out[, select], 1, sd)
out$lower <- out$effect - sds
out$upper <- out$effect + sds
out$rank <- nPRS:1

n <- dim(out)[1]
ggplot(data=out, aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=mean(out$effect), color='black', linetype='dashed') +
  geom_vline(xintercept=median(out$effect), color='blue', linetype='dashed') +
  theme_minimal()

```

- **Repeat the plot with only de PRS with $S_{R^2}$ larger then the median. Lines: in red and solid the 95% percentile; in red and dashed the 90%, 85%, 80% percentiles**
```{r}
sel <- out$Sum > median(out$effect)
ggplot(data=out[sel,], aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=quantile(out$Sum, prob=0.95), color='red', linetype='solid') +
  geom_vline(xintercept=quantile(out$Sum, prob=c(0.80, 0.85, 0.90)), color='red', linetype='dashed') +
  theme_minimal()
```

- **Create files with PRS with the sum of $R^2$ greater then the 95%, 90%, 85% and 80% percentile**
```{r}
sel_95 <- out$Sum > quantile(out$Sum, prob=0.95)
write.csv2(out[sel_95,-9], file="WExample1_precentil_95.csv")
sel_90 <- out$Sum > quantile(out$Sum, prob=0.90)
write.csv2(out[sel_90,-9], file="WExample1_precentil_90.csv")
sel_85 <- out$Sum > quantile(out$Sum, prob=0.85)
write.csv2(out[sel_85,-9], file="WExample1_precentil_85.csv")
sel_80 <- out$Sum > quantile(out$Sum, prob=0.80)
write.csv2(out[sel_80,-9], file="WExample1_precentil_80.csv")

```

## According to the obtained results, first PRS.1 is selected to analyze its association with the Trait

- **Step 2.1: Plots to check the possible interactions**
```{r}
ggplot(dat, aes(x=PRS.1, y=Trait)) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE)+
  facet_grid(Sex ~ Diagnostic, labeller=label_both)
```
## The plots suggest that the interaction between the PRS.1 and sex is relevant. Thus, we set the full model candidate (FM): $Trait \sim PRS + Sex + Diagnostic + PRS*Sex + PC1 +PC2$

- **Step2.2** Do parametric assumptions for the selected FM hold?
```{r}
#model
FM <- lm(Trait ~ PRS.1*Sex + Diagnostic + Age + PC1 + PC2, data=dat)
#qq-plot for normality 
plot(FM,2)   
```

```{r}
#Saphiro-Wilk test
shapiro.test(FM$residuals)
```

```{r}
#plot for variances
d <- fortify(FM)
ggplot(d,aes(x=.fitted, y=.stdresid, colour=Sex)) + geom_point() + geom_hline(yintercept=0, col="red")+
  facet_wrap(.~Sex)
```

```{r}
#Levene's test
leveneTest(.stdresid ~ Sex, data=d)
```

## It seems homocedasticity does no hold. Thus, we have two approaches to assess the possible association with PRS.1 and the Trait: try a Box-Cox transformation (**Step 4.2.1**) or perform a weighted permutation test (**Step 4.2.2**)

- **Step 4.2.1: Try a Box-Cox transformation of the dependent variable:**
```{r}
b <- boxcox(FM)
```
```{r}
# Exact lambda
lambda <- b$x[which.max(b$y)]
lambda
```
```{r}
dat$newTrait <- (dat$Trait ^ lambda - 1) / lambda
FM <- lm(newTrait ~ PRS.1*Sex + Diagnostic + Age + PC1 + PC2, data=dat)
```
```{r}
#Shapiro-Wilk test
shapiro.test(FM$residuals)
```

```{r}
#Levene's test
d <- fortify(FM)
leveneTest(.stdresid ~ Sex, data=d)
```
 
##In this case, the suggested Box-Cox transformation with λ = 0.667 does not solve the heteroscedasticity problem

- **Step 4.2.1: Weighted permutation test**
```{r}
NM <- lm(Trait ~ Sex + Diagnostic + Age + PC1+PC2, data=dat)
FM <- lm(Trait ~ PRS.1*Sex + Diagnostic + Age + PC1+PC2, data=dat)
outperm <- dR2(NullModel=NM, FullModel=FM, B=1000, seed=165,  weights=TRUE)
outperm
```

- **Last step: We move to the next PRS**

