
- **Title: Working example 2**
- **Output: html_document**
- **Reference: A guide to test association between Polygenic Risk Scores and psychological and psychiatric traits: practical examples**
- **Authors: Itziar Irigoien, Patricia Mas, Araceli Rosa, and Concepción Arenas**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
source("Functions.R")
library(ggplot2)
library(car)
library(multcomp)
```

## Working example 2

### In this example we simulate 10 PRSs, and a continuous trait, with gender, clinical diagnosis (with 2 categories), age, and two Principal Components as covariates. 

- **data reading**
```{r}
dat <- read.table("WExample2.csv", header=TRUE, sep=";", dec=",")
names(dat) #
```

- **do not forget to declare the categorical variables as factors**
```{r}
dat$Sex <- as.factor(dat$Sex)
dat$Diagnostic <- as.factor(dat$Diagnostic)
```

- **Step 1: Order  the PRSs**
```{r}
out <- orderR2(dat, yname="Trait", prsname = "PRS.")
head(out)
mainfilename <- "WExample2"
filename <- paste0(mainfilename, "_Ordered_PRS.csv")
write.csv2(out,file=filename)
```

- **Step 1: Plot the sum of coefficients of determination $S_{R^2}$. Lines: in blue the median; in black the mean**
```{r}
out <- data.frame(out) 
n <- dim(out)[1]
select <- grep("Model", names(out), value=FALSE)
out$effect <- out$Sum
sds <- apply(out[, select], 1, sd)
out$lower <- out$effect - sds
out$upper <- out$effect + sds
out$rank <- n:1


ggplot(data=out, aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=mean(out$effect), color='black', linetype='dashed') +
  geom_vline(xintercept=median(out$effect), color='blue', linetype='dashed') +
  theme_minimal()

```

- **Repeat the plot with only de PRS with $S_{R^2}$ larger then the median. Lines: in red and solid the 95% percentile; in red and dashed the 90%, 85%, 80% percentiles**
```{r}
sel <- out$Sum > median(out$effect)
ggplot(data=out[sel,], aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=quantile(out$Sum, prob=0.95), color='red', linetype='solid') +
  geom_vline(xintercept=quantile(out$Sum, prob=c(0.80, 0.85, 0.90)), color='red', linetype='dashed') +
  theme_minimal()
```

- **Create files with PRS with the sum of $R^2$ greater then the 95%, 90%, 85% and 80% percentile**
```{r}
sel_95 <- out$Sum > quantile(out$Sum, prob=0.95)
write.csv2(out[sel_95,-9], file=paste0(mainfilename, "_Percentile_95.csv"))
sel_90 <- out$Sum > quantile(out$Sum, prob=0.90)
write.csv2(out[sel_90,-9], file=paste0(mainfilename, "_Percentile_90.csv"))
sel_85 <- out$Sum > quantile(out$Sum, prob=0.85)
write.csv2(out[sel_85,-9], file=paste0(mainfilename, "_Percentile_85.csv"))
sel_80 <- out$Sum > quantile(out$Sum, prob=0.80)
write.csv2(out[sel_80,-9], file=paste0(mainfilename, "_Percentile_80.csv"))
```

### According to the obtained results, first PRS.5 is selected to analyse its association with the Trait.

- **Step 2.1: Plots to check the possible interactions**
```{r}
ggplot(dat, aes(x=PRS.5, y=Trait)) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE)+
  facet_grid(Sex ~ Diagnostic, labeller=label_both)
# Candidate FM Trait ~ PRS + Sex + Diagnostic + PRS + PRS*Diagnostic + PC1 +PC2
```

### The plots suggest that the interaction between the PRS.5 and the diagnostic is relevant. Thus, we set the full model candidate (FM): $Trait \sim PRS + Sex + Diagnostic + Sex + PRS \cdot Diagnostic + PC1 +PC2$

- **Step2.2: Do parametric assumptions for the selected FM hold?**
```{r}
#model
FM <- lm(Trait ~ PRS.5*Diagnostic + Sex + Age + PC1 + PC2, data=dat)
#qq-plot for normality 
plot(FM,2)   
```

### Lack of normality of residuals is suggested

```{r}
#Shapiro-Wilk test
shapiro.test(FM$residuals)
```

```{r}
#plot for variances
d <- fortify(FM)
ggplot(d,aes(x=.fitted, y=.stdresid, colour=Diagnostic)) + geom_point() + geom_hline(yintercept=0, col="red")+
  facet_wrap(.~Diagnostic)
```

```{r}
#Levene's test
leveneTest(.stdresid ~ Diagnostic, data=d)
```

### It seems that there is lack of normality of residual but linearity and homocedasticity assumptions hold. Thus, we have two approaches to assess the possible association with PRS.5 and the Trait: try a transformation (**Step 4.1.1**) or perform a permutation test (**Step 4.1.2**)

- **Step 4.1.1: Try a transformation of the dependent variable:**
```{r}
dat$logTrait <- log(dat$Trait)
FM <- lm(logTrait ~ PRS.5*Diagnostic + Sex + Age + PC1 + PC2, data=dat)
#qq-plot for normality 
plot(FM,2)
```
```{r}
#Shapiro-Wilk test
shapiro.test(FM$residuals)
```

### Since it is suggested the transformation offered a solution for the lack of normality, the model we build is given by:

```{r}
summary(FM)
```

### The results show that PRS.5 is related with the log(Trait) in the following way:

\begin{itemize}
    \item $\widehat{log(Trait)} = 2.585 + 0.00089\times PRS.5 + 0.041\times Sex - 0.0006\times Age + 0.197\times PC1 - 0.0596\times PC2$ if Diagnostic = 0.
    \item $\widehat{log(Trait)} = (2.585-20.052) + (0.00089-0.26068)\times PRS.5 +  0.041\times Sex + - 0.0006\times Age  + 0.197\times PC1 - 0.0596\times PC2$ if Diagnostic = 1.
  \end{itemize}

- **Step 3: Check whether the respective PRS coefficients under each group are significant or not**
```{r}
summary(glht(FM, "PRS.5 = 0"))
summary(glht(FM, "PRS.5  + PRS.5:Diagnostic1 = 0"))
```

### That means that for Diagnostic 1, the association is negative (coeff = -0.25979, $p$-value= 0.004487 (*)); for Diagnostic 0, there is not association (coeff = 0.00089, $p$-value=0.9868). Given this transformation, the increase of one unit in PRS.5, makes the ration $\widehat{log(Trait)}|PRS.5 = 1/\widehat{log(Trait)|PRS.5=0} = exp(-0.25979)=0.771 < 1$, thus, gives a mean decrease of about 22.9\%


- **Step 4.1.2: Perform a permutation test to assess the association**
```{r}
NM <- lm(Trait ~ Diagnostic + Sex + Age + PC1 + PC2, data=dat)
FM <- lm(Trait ~ PRS.5*Diagnostic + Sex + Age + PC1 + PC2, data=dat)
outperm <- dR2(NullModel=NM, FullModel=FM, B=1000, seed=165)
outperm
```
### We observe an increase of 0.108 in the coefficient of determination when the PRS.5 is included in the model and the permutation test indicates it is significant.

- **Last step: We move to the next PRS**

