
- **Title: Working example 1**
- **Output: html_document**
- **Reference: Tutorial: a guide to test evidence of association between a Polygenic Risk Score and a trait**
- **Authors: Itziar Irigoien, Patricia Mas, Araceli Rosa, and Concepción Arenas**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
source("Functions.R")
library(ggplot2)
library(car)
library(multcomp)
```


## Working example 1

## In this example we simulate 9 PRSs, and a continuous trait, with gender, clinical diagnosis (with 2 categories), age, and two Principal Components as covariates

- **data reading**
```{r}
dat <- read.table("WExample1.csv", header=TRUE, sep=";", dec=",")
names(dat) #
```

- **do not forget to declare the categorical variables as factors**
```{r}
dat$Sex <- as.factor(dat$Sex)
dat$Diagnostic <- as.factor(dat$Diagnostic)
```

- **Step 1: Order  the PRSs**
```{r}
out <- orderR2(dat, yname="Trait", prsname = "PRS.")
head(out)
write.csv2(out,file="WExample1_Ordered_PRS.csv")
```

- **Step 1: Plot the sum of coefficients of determination $S_{R^2}$. Lines: in blue the median; in black the mean**
```{r}
out <- data.frame(out) 
nPRS <- dim(out)[1]
select <- grep("Model", names(out), value=FALSE)
out$effect <- out$Sum
sds <- apply(out[, select], 1, sd)
out$lower <- out$effect - sds
out$upper <- out$effect + sds
out$rank <- nPRS:1

n <- dim(out)[1]
ggplot(data=out, aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=mean(out$effect), color='black', linetype='dashed') +
  geom_vline(xintercept=median(out$effect), color='blue', linetype='dashed') +
  theme_minimal()

```

- **Repeat the plot with only de PRS with $S_{R^2}$ larger then the median. Lines: in red and solid the 95% percentile; in red and dashed the 90%, 85%, 80% percentiles**
```{r}
sel <- out$Sum > median(out$effect)
ggplot(data=out[sel,], aes(y=rank, x=effect, xmin=lower, xmax=upper)) +
  geom_point() +
  geom_errorbarh(height=.1) +
  scale_y_continuous(name=NULL, breaks= n:1, labels=row.names(out), position="right") +
  labs(title='', x='Sum R^2', y = 'PRS') +
  geom_vline(xintercept=quantile(out$Sum, prob=0.95), color='red', linetype='solid') +
  geom_vline(xintercept=quantile(out$Sum, prob=c(0.80, 0.85, 0.90)), color='red', linetype='dashed') +
  theme_minimal()
```

- **Create files with PRS with the sum of $R^2$ greater then the 95%, 90%, 85% and 80% percentile**
```{r}
sel_95 <- out$Sum > quantile(out$Sum, prob=0.95)
write.csv2(out[sel_95,-9], file="WExample1_precentil_95.csv")
sel_90 <- out$Sum > quantile(out$Sum, prob=0.90)
write.csv2(out[sel_90,-9], file="WExample1_precentil_90.csv")
sel_85 <- out$Sum > quantile(out$Sum, prob=0.85)
write.csv2(out[sel_85,-9], file="WExample1_precentil_85.csv")
sel_80 <- out$Sum > quantile(out$Sum, prob=0.80)
write.csv2(out[sel_80,-9], file="WExample1_precentil_80.csv")

```

## According to the obtained results, first, PRS.4 is selected to analyse its possible association with the Trait

- **Step 2.1: Plots to check the possible interactions**
```{r}
# First candidate PRS.4
# Plot it
ggplot(dat, aes(x=PRS.4, y=Trait)) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE)+
  facet_grid(Sex ~ Diagnostic, labeller=label_both)
```

## The plots suggest that the interaction between the PRS.4 and the diagnostic is relevant. Thus, we set the full model candidate (FM): $Trait \sim PRS + Sex + Diagnostic + PRS*Diagnostic + PC1 +PC2$


- **Step2.2: Do parametric assumptions for the selected FM hold?**
```{r}
#model
FM <- lm(Trait ~ PRS.4*Diagnostic + Sex + Age + PC1 + PC2, data=dat)
#qq-plot for normality 
plot(FM,2)   
```

```{r}
#Saphiro-Wilk test
shapiro.test(FM$residuals)
```

```{r}
#plot for variances
d <- fortify(FM)
ggplot(d,aes(x=.fitted, y=.stdresid, colour=Diagnostic)) + geom_point() + geom_hline(yintercept=0, col="red")+
  facet_wrap(.~Diagnostic)
```

```{r}
#Levene's test
leveneTest(.stdresid ~ Diagnostic, data=d)
```

## It seems all assumptions fulfill, therefore, we move to *Step 3*

- **Step 3: Check the coefficients of the model**
```{r}
summary(FM)
```
## The results show that PRS.4 is related with the Trait in the following way:

\begin{itemize}
    \itemif Diagnostic = 0, $\widehat{Trait} = 14.816 + -1.041\, PRS.4 + 1.3\,Sex -0.071\,Age + 0.927\,PC1 + 8.033\,PC2$ 
    \item if Diagnostic = 1, $\widehat{Trait}$ = (14.816 -8.321) + (-1.041+15.022)\,PRS.4 + 1.3\,Sex -0.071\,Age + 0.927\,PC1 + 8.033\,PC2 

\end{itemize}


- **Step 3: Check whether the respective PRS coefficients under each group are significant or not**
```{r}
summary(glht(FM, "PRS.4 = 0"))
summary(glht(FM, "PRS.4  + PRS.4:Diagnostic1 = 0"))
```

## That means that for Diagnostic group 0 there is no relevant association (coeff = -1.041), but for Diagnostic group 1, the association is strong, positive (coeff = 13.980), and significant (p-value = 0.014)


- **Last step: We move to the next PRS**

